---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Lindsay Stall"
date: "1/31/23"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r}
library(usethis)
use_git_config(user.name = "Lindsay Stall", 
               user.email = "stallm21@wfu.edu")
```

```{r}
staff <- read_csv("data/instructional-staff.csv")
```
```{r}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```

```{r}
staff_long
```
### Exercise 1

```{r}
staff_long %>%
  ggplot(aes(x = year,
             y = value,
             group = faculty_type,
             color = faculty_type,)) +
  labs(
    x = "Years",
    y = "Faculty type",
    title = "Faculty Type by Year"
  ) +
  geom_line()
```

### Exercise 2

I would create 2 groups: one for part-time and one for full time to make the rates of employment between the groups easier to compare. 

### Exercise 3

```{r}
fisheries <- read_csv("data/fisheries.csv")
```
```{r}
fisheries_long <- fisheries %>%
  pivot_longer(cols = -country, names_to = "totals") %>%
  filter(totals %in% c("capture", "aquaculture")) %>%
  mutate(value = as.numeric(value))
```

```{r}
fisheries_long %>%
  filter(country %in% c("China", "Indonesia", "India", "United States", "Russia", "Peru", "Vietnam")) %>%
  ggplot(aes(fill = totals, y=value, x=country)) +
  labs(
    x = "Countries",
    y = "Tons of Fish",
    title = "Tons of Fish Produced by Top Ten Fish-Producing Countries"
  ) +
  geom_bar(position="stack", stat="identity")
```


### Exercise 3
```{r}
library(tidyverse) 
library(mosaicData) 
```

```{r}
data(Whickham)
```

```{r}
?Whickham
```

```{r}
view(Whickham)
```

What type of study do you think these data come from: observational or experiment? Why?
Observational because it is a survey and the outcomes were dead and alive.

How many observations are in this dataset? What does each observation represent?
1314 observation, each observation is a person surveyed.

How many variables are in this dataset? What type of variable is each? Display each variable using an appropriate visualization.
3 variables:
outcome survival status after 20 years: a factor with levels Alive Dead
smoker smoking status at baseline: a factor with levels No Yes
age age (in years) at the time of the first survey

What would you expect the relationship between smoking status and health outcome to be?
"Yes" status would lead to greater "dead" outcomes (esp at younger ages)

### Exercise 3

```{r}
Whickham %>%
  count(smoker, outcome)
```

```{r}
ggplot(Whickham, aes(fill = smoker, x = outcome))+
geom_bar()
```
68% of Non smokers are alive
76% of smokers are alive
This was not what I expected, as I was expecting more dead smokers.

### Exercise 3

```{r}
Whickham <- Whickham %>%
  mutate(age_cat = case_when(
    age <= 44 ~ "18-44",
    age > 44 & age <= 64 ~ "45-64",
    age > 64 ~ "65+"
))
```


```{r}
Whickham %>%
  count(smoker, age_cat, outcome)
```
```{r}
ggplot(Whickham, aes(fill = smoker, x = outcome))+
facet_wrap(~age_cat)+geom_bar()
```

The proportion of alive smokers in the 65+ category is very small. In the 18-44 and 45-64 categories, there is a higher proportion dead people that are smokers than non-smokers. 
